Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-18 jso.qmd
edits in sec5

In addition, to assess the performance of JSO across various PP problems, two different
data shapes, pipe and a sine wave, are considered in various dimensional spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimisation success. The number of jellies marginally affects success, but index smoothness, and increasing the number of tries do not."
#| fig-pos: t
library(gtsummary)
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    #long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

@@

Squintability and smoothness are likely dependent, making it unnecessary to include both as independent factors in the model. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness.

>>>>

In addition, to assess the performance of JSO across various PP problems, two different data shapes, pipe and a sine wave, are considered in various dimensional spaces using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. This study does not consider dimensions lower than four, as exploring the search space in such cases is relatively straightforward. Similarly, extremely high-dimensional cases with complex data structures, where no method can reliably and consistenly uncover the underlying structure due to the curse of dimensionality, are also excluded, as they fall outside the scope of this study. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

@@

```{r}
#| tbl-pos: H
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimisation success. The number of jellies marginally affects success, but index smoothness, and increasing the number of tries do not."
#| fig-pos: t
library(gtsummary)
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    #long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

@@

The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness.

### 2025-02-18 jso.tex
edits in sec5

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, toprule=.15mm, colback=white, leftrule=.75mm, arc=.35mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, rightrule=.15mm, toptitle=1mm, bottomrule=.15mm, breakable, titlerule=0mm, colframe=quarto-callout-note-color-frame]

@@

In addition, to assess the performance of JSO across various PP
problems, two different data shapes, pipe and a sine wave, are
considered in various dimensional spaces using six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, under varied JSO
hyper-parameters. This results in a total of 76 scenarios, comprising of
30 computed on the pipe data and 46 on the sine-wave data. A list of
combinations of the data shape, dimensions and PP indexes that are
considered in this study is provided in
Table~\ref{tbl-smoothness-squintability}. Again, JSO is run 50 times for
each scenario to calculate the success rate for each hyper-parameter
combination.

@@

@@

Squintability and smoothness are likely dependent, making it unnecessary
to include both as independent factors in the model. Intuitively, a
highly smooth search space exhibits gradual changes, making it easier to
perceive the overall structure from a distance, thereby increasing the
squint angle. Table~\ref{tbl-joint-output} compare the model fits when
squintability and smoothness are included separately. Given that the
null deviance remains the same across models, the model with
squintability yields a smaller residual deviance and lower AIC and BIC
values, indicating a better fit. This suggests that squintability
captures the relevant variation more effectively than smoothness.

>>>>

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, titlerule=0mm, opacityback=0, coltitle=black, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, arc=.35mm, breakable, toptitle=1mm, left=2mm, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm]

@@

In addition, to assess the performance of JSO across various PP
problems, two different data shapes, pipe and a sine wave, are
considered in various dimensional spaces using six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, under varied JSO
hyper-parameters. This results in a total of 76 scenarios, comprising of
30 computed on the pipe data and 46 on the sine-wave data. This study
does not consider dimensions lower than four, as exploring the search
space in such cases is relatively straightforward. Similarly, extremely
high-dimensional cases with complex data structures, where no method can
reliably and consistenly uncover the underlying structure due to the
curse of dimensionality, are also excluded, as they fall outside the
scope of this study. A list of combinations of the data shape,
dimensions and PP indexes that are considered in this study is provided
in Table~\ref{tbl-smoothness-squintability}. Again, JSO is run 50 times
for each scenario to calculate the success rate for each hyper-parameter
combination.

@@

\begin{table}[H]

@@

The simulation result suggests that squintability and smoothness are
likely dependent. Intuitively, a highly smooth search space exhibits
gradual changes, making it easier to perceive the overall structure from
a distance, thereby increasing the squint angle.
Table~\ref{tbl-joint-output} compare the model fits when squintability
and smoothness are included separately. Given that the null deviance
remains the same across models, the model with squintability yields a
smaller residual deviance and lower AIC and BIC values, indicating a
better fit. This suggests that squintability captures the relevant
variation more effectively than smoothness.
